---
title: "conditional programming"
author: "Vinay Swamy"
date: "9/7/2020"
output: ioslides_presentation
---

<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>
```{r, include=F}
library(tidyverse)
```

## Control flow and conditional programming

- There are a couple more programming concepts we need to cover.
- Some times when solving a problem,  we want to account for two(or more) different outcomes, but at any given time we only want one of them to happens. This is called conditional programming

## `if` statement

- controlling code is done using `if` statements. Consider the following examples
```{r}
is_greater_than_five <- function(number){
    if(number > 05){
        return('Greater than five')
    } 
   
    return('the number is less than five')

}
```

## `if` statement

```{r}
is_greater_than_five(10)
```
```{r}
is_greater_than_five(4)
```

## `if` statement

- this is the general structure of an `if` statement
```
if(LOGICAL_VALUE){
    Everything within these brackets 
    is executed
}

This is outside the if statement.
```

## `logical` values

- the `logical` is the final data type in R.
- There are 2 possible values a `logical`: `TRUE` and `FALSE`
- Generally, `logical`s are generated when we want to check whether a certain condition is met using a logical operator or function
- you are likely already familiar with two logical operators, `>` and `<` . These function just like they do in math

```{r}
x <- 5
x>3
```

## `logical` values

```{r}
x <- 2
x>3
```

## `logical` values
- under the hood, TRUE and FALSE are represented by the numbers 1 and zero
```{r}
tf <- c(TRUE, FALSE)
as.numeric(tf)
```
- using `sum` with a `logical` vector is a good way to count the number elements that match a certain condition

## logical operations over vectors
- all logical operators are vecotorized, meaning they can apply they can be applied directly to all elements in a vector

```{r}
x <- c(1, 9, 6, 2, 3, 6)
x>5

```

## logical operations over vectors
vectors of the same length can also be compared, but this creates an element wise comparision and returns a vector
```{r}
x <- c(1, 9, 6, 2, 3, 6)
y <- c(2, 4, 10, 5, -4)
comp <-  x>y
comp
```
```{r}
str(comp)
```

## logical operations over vectors
- to compare two vectors, us,gvs match each other; `any` to see if at least one elements between the two match

```{r}
x <- c(1, 9, 6, 2, 3, 6)
y <- c(2, 4, 10, 5, 1, -4)
all(x==y)
```

## the `==` operator
- the `==` is the equality opeator, and checks whether two values are **exactly** the same 

```{r}
5==5
```
```{r}
5  == 4.9999
```
```{r}
x <- c(1, 9, 6, 2, 3, 6)
x==6
```

## the `!` operator

-this operator inverts an existing logical vector; all trues become false, and all falses become true
- `T` can be used for `TRUE` and `F` can be used for `FALSE`
```{r}
x <- c(T,F,T,T,F)
x
```
```{r}
!x
```

## the `%in%` operator

- the `%in%` operator allows you to check whether elements in one vector are in another


```{r}
vec_1 <- c('A', 'B', 'C', 'D', 'E', 'F')
vec_2 <- c('A', 'b', 'C', 'd', 'E')

vec_1 %in% vec_2
```
```{r}
vec_1[vec_1 %in% vec_2]
```
- Note that because it check whether elements of the *left* are in the *right*, a vector equal to the length of the left is returned


## the `else` block
- when you're using an `if` statement, generally if the coniditon for the `if` statement is not met, you somthing else to happen

```{r}

weather_report <- function(is_sunny){
    if(is_sunny){
        print('it is a nice day out')
        
    }else{
        print('maybe stay inside today')
    }
}
weather_report(is_sunny = T)
```
```{r}
weather_report(is_sunny = F)
```
-Its important to note that these two blocks will *never* run at the same time; one or the other will always occur

## the `else` block

-By just looking at the code below, what will the outputs of `weather_report_if_else(T)` and `weather_report_if(T)` be?
```{r}
weather_report_if_else <- function(is_sunny){
    if(is_sunny){
        print('it is a nice day out')
        
    }else{
        print('maybe stay inside today')
    }
}

weather_report_if <- function(is_sunny){
    if(is_sunny){
        print('it is a nice day out')
        
    }
    
    print('maybe stay inside today')
    
}
```

## the `else` block

```{r}
weather_report_if_else(T)
```

## the `else` block

```{r}
 weather_report_if(F)
```

## the `else` block
- This is an example of "leaky" control flow, where the two outcomes are not proerly separated; This is why its best practice to always inlcude an `else` block as part of an if statement

## accounting for multiple conditions
- some times, we want to account for more than just 2 outcomes. This is done by adding `else if` blocks 

```{r}
weather_report <- function(is_sunny, is_raining){
    if(is_sunny){
        print('it is a nice day out')
        
    }else if(is_raining){
        print('Bring an umbrella')
    }
    else{
        print('maybe stay inside today')
    }
}


weather_report(is_sunny = F, is_raining = T)
```

## accounting for multiple conditions

```{r}
weather_report(is_sunny = F, is_raining = F)
```
## accounting for multiple conditions

```{r}
weather_report(is_sunny = T, is_raining = F)
```

## accounting for multiple conditions
- when using `else if` blocks, its important to remember if multiple conditions are met, only the *first* block is executed

```{r}
weather_report(is_sunny = T, is_raining = T)
```


## logical indexing

- one of the most pratical uses of logical operators is for logical indexing.
-  in logical indexing we select elements that meet a certain condition

```{r}
vec <- c(1, -2, 4, -5, 8, -3)
keep_above_zero <- vec >0
vec[keep_above_zero]
```

## logical indexing
- this can be directly done within the indexer
```{r}
vec[vec > 0]
```
-logical indexing can be used in any situation normal indexing can

## vectorized `if` statements

- the `ifelse` function is a good way to apply an if-else condition with a binary outcome over a logical vector 
- for example for each element in a vector, if the number is greater than 0 , set it to `Inf`, and if its below 0, set it to `-Inf`

```{r}
vec <- c(1,30000,-1, 354, -4)
vec_greater_than_0 <- vec > 0
ifelse(test = vec_greater_than_0, yes =Inf, no = -Inf  )# yes and no correspond to the true and false case respectively

```

## useful base R functions when working with logical vectors
- We're going to cover some of the more useful functions within the base R that use logical vectors

## Replace
- `replace` allows you to change a vector based on a logical index
```{r}
data(mpg)
head(mpg)
```

## Replace
```{r}
mpg$trans <- replace(x = mpg$trans, list = mpg$trans == 'auto(l5)', 'auto L-5')
head(mpg)
```

## grepl
- `grepl` searches a character vector for a pattern, and returns TRUE/FALSE where the pattern is matched. 

```{r}
nrow(mpg)
```
## grepl

```{r}
audi <- grepl(pattern = 'audi', mpg$manufacturer, fixed = T)
nrow(mpg[audi, ])
```
## grepl

```{r}
head(mpg[audi, ])
```

## grepl
- by default, `grepl` uses regular expressions(regex) to match patterns. They are fairly complicated, and we're not gonna learn about them in this course, but highly recommend doing so on your own if you want to keep using R.
- to avoid using regex, always use the `fixed = T` flag





