---
title: "R Notebook"
output: html_notebook
---

# Practice with `ggplot2`

We're going to go through several of the common geoms in ggplot. If at anytime you are stuck, try using the `?` to get the documentation for a function , or googling the problem. 

We're making plots in Rmarkdown; plots will display below the chunk once you run the chunk with the play button. 

## `geom_point`

First, we'll be going through the geom_point function. We'll be using the `CO2` dataframe, The CO2 data frame has 84 rows and 5 columns of data from an experiment on the cold tolerance of the grass species Echinochloa crus-galli.

```{r}
data("CO2")
CO2
```

load the ggplot2 library, and make a scatter plot using `conc` for the x axis and `uptake` for the y axis.

```{r}


```


Now change all the points to the color red

```{r}

```


Change the color of the points to color by the `Type` column

```{r}

```

Change the color of the points to color by the `Plant` column` column

```{r}
x
```

note that this coloring is a continuous coloring, ie a uniform transition between shades of color. Continous data types like `numeric` or `ordered.factor` are automatically given continous scales, whereas categorical types like `character` or `factor` are colored with a discrete scales. We'll talk more about this when we discuss colors 

change the `Plant` column to a character and re create the plot above to see the change in color

```{r}

```

- now change the shape of the points based on the `Treatment` column.  A combination of shapes and colors is a good way to show two different variables

```{r}

```

the `alpha` value of a `geom` sets how transparent it is, with 1 being fully visible, and 0 being fully transparent

set the `alpha` value for all points to .75, then .5, then .25 to see how the plot changes

```{r}

```


## line graphs 

We'll be working with the `Indometh` data set, which has data related  Pharmacokinetics of Indomethacin
```{r}
#run me
data("Indometh")
Indometh
```


make a basic line graph comparing `time` and `conc` from the `Indometh` data 

```{r}

```


Now make seperate lines using a different color for `different subject`.
```{r}

```

instead of connecting the data, use fitted line to make a similar plot as above, with separate lines for each subject. Try out a couple different smoothing methods

## Bargraphs 
we're going to practice making bar graphs use the `mpg` data set which contains Fuel economy data from 1999 and 2008 for 38 popular models of car

```{r}
data(mpg)
mpg
```

make a bargraph for  the `manufacturer` column

```{r}

```
- dpending on the size of your screen, the labels on the x-axis maybe overlapping. We'll learn how to fix this in a later lesson

plot a stacked bar class using `class` as the coloring variable. 

```{r}

```

now plot a bargraph where the Y axis shows the `hwy` column, and `class` for the x axis

```{r}

```

conver the `cyl` column to a `character`, and then make a dodged barplot using `cyl` as the coloring variable 

```{r}

```

## Boxplots 
 
We'll be working with the ToothGrowth data set, which has data related to the effect of Vitamin C on Tooth Growth in Guinea Pigs
 
```{r}
data("ToothGrowth")
ToothGrowth
```

use a boxplot to visualize the length(`len`) distributions for the different treatment conditions(`supp`). Use the `?` command if you get stuck

```{r}

```

add `theme_minimal` to this plot if you have not done so; note the changes 

```{r}

```


change the shading of the box to fill the box based on treatment

```{r}

```

change the outline of the box based on the treatment

```{r}

```


change the width of all boxplots 
```{r}

```



Now change the width of the lines within the boxplot; which parameter to change might not be clear from the documentation, so this may take some trial error
```{r}

```

## Visualization distributions 

Below is some fake data about the ages of patients in various part of a hospital. We are trying to visualize whether the distributions of ages in each region are different

```{r}
mu <- 2
si <- 0.6
hospital_data <- data.frame( ICU  =  c(rnorm(1000, -mu, si), rnorm(1000, mu, si)), 
                         ER  = runif(1000, -4, 4),
                         Clinic   =  rnorm(2000, 0, si*4)) %>% 
    tidyr::pivot_longer(cols = ICU:Clinic, 'Hospital_location') %>% rename(age = 'value')

```


first, lets plot a boxplot; make multiple boxplots shaded by location

```{r}

```

now plot make violin plots for each location; take note of the distributions 

```{r}

```


now plot histograms, again shaded by location

```{r}

```

`geom_freqpoly` is  drop-in replacement for `geom_histogram`: instead of drawing bar's for the count in each bin, it draws points at each count, and then draws a line through each point. Try it out with the above data, tweaking the coloring if necessary


```{r}

```


## Practice with themes 

We're going to work with one of the plots we made earlier 


```{r}
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer, fill = class)) 

```


Assign descriptive axis labels and a plot title to the plot

```{r}

```


fix the labels similar to how it was done in the example

```{r}

```


change the y axis title so it is parallel to the x axis title 

```{r}

```

change the font of the title to AvantGarde, the color to red, and make it italizced


```{r}

```


## practice with colors 



