---
title: "Visualization with R day 2"
author: "Vinay Swamy"
date: "9/7/2020"
output: ioslides_presentation
---

## Review 

## Data Types
- We have covered 3 data types in R. What are they, and what is each on used for 


## Data Types
- `numeric` - used for numerical data ; `character` used for strings/text

## List vs Vector

- What is the major difference between a list and a vector

## List vs Vector

- vectors must be a single data type; lists can contain multiple data types, and can contain multiple data stuctures, ie lists of lists, lists of dataframes 


## Dataframes
- what type of data are dataframes used for 

## Dataframes
- tabular data like spreadsheets or csv's 

## Indexing

- how does indexing work for a vector
- how does indexing work for a  list
- how does indexing work for a  dataframe?

## Indexing

- vector
    - `vec[i]`, for the ith  element in a vector
    - `vec[i:j]` to select all elements between i and j 
- list
    - `lst[[i]]`, for the ith  element in a list
    - `lst[i:j]`, to select all elements between i and j 
- dataframe 
    - `df[, i]` to select column number i; `df[i, ]
    - `df[i,]` to select row number i; 
    - `df[i:j, ]` / `df[,i:j]` for selecting a range of rows/columns 
    
- if an object has names, values can also be accessed by their name, instead of their numerical index

##  ggplot

- what are the three major components of a `ggplot`?

##  ggplot

-  base `ggplot` layer, `geom` layer, `theme` layer


## Control flow and conditional programming

- There are a couple more programming concepts we need to cover.
- Some times when solving a problem,  we want to account for two(or more) different outcomes, but at any given time we only want one of them to happens. This is called conditional programming

## `if` statements 

- controlling code is done using if statements. Consider the following examples
```{r}
is_greater_than_five <- function(number){
    if(number > 05){
        return('Greater than five')
    } 
   
    return('the number is less than five')

}
```
```{r}
is_greater_than_five(10)
```
```{r}
is_greater_than_five(4)
```


## `if` statements

- this is the  general structure of an if statement

```
if(LOGICAL_VALUE){
    Everything within these brackets 
    is executed
}

This is outside the If statement.
```

## `logical` values

- the `logical` is the final data type in R.
- There are 2 possible values a `logical`: `TRUE` and `FALSE`
- Generally, `logical`s are generated when we want to check whether a certain condition is met using a logical operator or function
- you are likely already familiar with two logical operators, `>` and `<` . These function just like they do in math


```{r}
x <- 5
x>3
```

```{r}
x <- 2
x>3
```

- all logical operators are vecotorized, meaning they can apply they can be applied directly to all elements in a vector

```{r}
x <- c(1, 9, 6, 2, 3, 6)
x>5

```

vectors of the same length can also be compared, but this creates an element wise comparision and returns a vector
```{r}
x <- c(1, 9, 6, 2, 3, 6)
y <- c(2, 4, 10, 5, -4)
comp <-  x>y
comp
```
```{r}
str(comp)
```

- to compare two vectors, us,gvs match each other; `any` to see if at least one elements between the two match


```{r}
x <- c(1, 9, 6, 2, 3, 6)
y <- c(2, 4, 10, 5, 1, -4)
all(x==y)
```






## the `==` operator
- the `==` is the equality opeator, and checks whether two values are **exactly** the same 

```{r}
5==5
```
```{r}
5  == 4.9999
```
```{r}
x <- c(1, 9, 6, 2, 3, 6)
x==6
```

## the `!` operator

-this operator inverts an existing logical vector; all trues become false, and all falses become true
- `T` can be used for `TRUE` and `F` can be used for `FALSE`
```{r}
x <- c(T,F,T,T,F)
x
```
```{r}
!x
```

## the `%in%` operator

- the `%in%` operator allows you to check whether elements in one vector are in another


```{r}
vec_1 <- c('A', 'B', 'C', 'D', 'E', 'F')
vec_2 <- c('A', 'b', 'C', 'd', 'E')

vec_1 %in% vec_2
```
```{r}
vec_1[vec_1 %in% vec_2]
```
- Note that because it check whether elements of the *left* are in the *right*, a vector equal to the length of the left is returned


## the `else` block
- when you're using an `if` statement, generally if the coniditon for the `if` statement is not met, you somthing else to happen

```{r}

weather_report <- function(is_sunny){
    if(is_sunny){
        print('it is a nice day out')
        
    }else{
        print('maybe stay inside today')
    }
}
weather_report(is_sunny = T)
```
```{r}
weather_report(is_sunny = F)
```
-Its important to note that these two blocks will *never* run at the same time; one or the other will always occur


-By just looking at the code below, what will the outputs of `weather_report_if_else(T)` and `weather_report_if(T)` be?
```{r}
weather_report_if_else <- function(is_sunny){
    if(is_sunny){
        print('it is a nice day out')
        
    }else{
        print('maybe stay inside today')
    }
}

weather_report_if <- function(is_sunny){
    if(is_sunny){
        print('it is a nice day out')
        
    }
    
    print('maybe stay inside today')
    
}
```


```{r}
weather_report_if_else(T)
```

```{r}
 weather_report_if(T)
```

- This is an example of "leaky" control flow, where the two outcomes are not proerly separated; This is why its best practice to always inlcude an `else` block as part of an if statement


## accounting for multiple conditions
- some times, we want to account for more than just 2 outcomes. This is done by adding `else if` blocks 


```{r}
weather_report <- function(is_sunny, is_raining){
    if(is_sunny){
        print('it is a nice day out')
        
    }else if(is_raining){
        print('Bring an umbrella')
    }
    else{
        print('maybe stay inside today')
    }
}


weather_report(is_sunny = F, is_raining = T)
```
```{r}
weather_report(is_sunny = F, is_raining = F)
```
```{r}
weather_report(is_sunny = T, is_raining = F)
```

when using `else if` blocks, its important to remember if multiple conditions are met, only the *first* block is executed

```{r}
weather_report(is_sunny = T, is_raining = T)
```

## accounting for multiple conditions with logical operators
- another way of combining multiple condition is by using the comparison operators, `|` and `&`



## logical indexing

- one of the most pratical uses of logical operators is for logical indexing.
-  in logical indexing we select elements that meet a certain condition

```{r}
vec <- c(1, -2, 4, -5, 8, -3)
keep_above_zero <- vec >0
vec[keep_above_zero]
```
- this can be directly done within the indexer
```{r}
vec[vec > 0]
```

-logical indexing can be used in any situation normal indexing can

## vectorized `if` statements

- the `ifelse` function is a good way to apply an if-else condition with a binary outcome over a logical vector 
- for example for each element in a vector, if the number is greater than 0 , set it to `Inf`, and if its below 0, set it to `-Inf`

```{r}
vec <- c(1,30000,-1, 354, -4)
vec_greater_than_0 <- vec > 0
ifelse(test = vec_greater_than_0, yes =Inf, no = -Inf  )# yes and no correspond to the true and false case respectively

```











