---
title: "Day_1_review_and_conditional_programming_slides"
author: "Vinay Swamy"
date: "9/7/2020"
output: ioslides_presentation
---


## Control flow and conditional programming

- before we get to cleaning data, there are a couple more programming concepts we need to cover.
- Some times when solving a problem,  we want to account for two(or more) different outcomes, but at any given time we only want one of them to happens. This is called conditional programming


## `if` statements 

- controlling code is done using if statements. Consider the following examples
```{r}
is_greater_than_five <- function(number){
    if(number > 05){
        return('Greater than five')
    } 
   
    return('the number is less than five')

}
```
```{r}
is_greater_than_five(10)
```
```{r}
is_greater_than_five(4)
```

s
## `if` statements

- this is the  general structure of an if statement

```
if(LOGICAL_VALUE){
    Everything within these brackets 
    is executed
}

This is outside the If statement.
```

## `logical` values

- the `logical` is the final data type in R.
- There are 2 possible values a `logical`: `TRUE` and `FALSE`
- Generally, `logical`s are generated when we want to check whether a certain condition is met using a logical operator or function
- you are likely already familiar with two logical operators, `>` and `<` . These function just like they do in math


```{r}
x <- 5
x>3
```

```{r}
x <- 2
x>3
```

- all logical operators are vecotorized, meaning they can apply they can be applied directly to all elements in a vector

```{r}
x <- c(1, 9, 6, 2, 3, 6)
x>5

```

vecotors of the same length can also be compared, but this creates an element wise comparision and returns a vector
```{r}
x <- c(1, 9, 6, 2, 3, 6)
y <- c(2, 4, 10, 5, 1, -4)
comp <-  x>y
comp
```
```{r}
str(comp)
```

## the `==` operator
- the `==` is the equality opeator, and checks whether two values are **exactly** the same 

```{r}
5==5
```
```{r}
5  == 4.9999
```
```{r}
x <- c(1, 9, 6, 2, 3, 6)
x==6
```

## the `!` operator

-this operator inverts an existing logical vector; all trues become false, and all falses become true
- `T` can be used for `TRUE` and `F` can be used for `FALSE`
```{r}
x <- c(T,F,T,T,F)
x
```
```{r}
!x
```

