---
title: "Review + tidying data"
output: html_notebook
---

## conditional programming practice + review 

We'll be using some sample data from New York Air Quality Measurements. Notice how some of the data is missing

```{r}
data("airquality")
airquality
```

the `is.na` function takes a vector as input and returns `TRUE` when there is a `NA`, and `FALSE` otherwise. Use the `is.na` function on the `Ozone` column and print the results

```{r}

```


 try to calculate the mean for the `Ozone` column 

```{r}

```


For the `Ozone` column values that are NOT(!) `NA`, and calculate the mean and store it in a variable

```{r}

```


select the `NA` values in the `Ozone` to the mean value we calculated earlier

```{r}

```


use the `%in%` operator to select values from the `Month` column equal to 6 or 7. Store this in a variable

```{r}

```

use the above logical indexing vector to select rows from the `airquality` dataframe; keep all columns 

```{r}

```

use the `ifelse` function to replace the numbers corresponding to each month with their actual names () 6 -> "June", 7->"July")

```{r}

```

make a boxplot for the `Ozone` levels for each month, coloring boxplots by month, including descriptive name for the axes and a title for the plot.

```{r}

```






