---
title: "Manipulating data"
author: "Vinay Swamy"
date: "9/4/2020"
output: ioslides_presentation
---


## Data usually does not come with a bow on top

- almost all real world data won't be ready for plotting; usually steps to *clean* and *tidy* the data are required
- clean data as all the data with the right data types, ie numbers are `numeric`, strings are `characters`
- tidy data has data organized such that each complete row is a unique observation.
- thankfully there are quite a bit of resources for doing this within R


## Libraries in R

- up until now we either been using functions included within the base R environment and if function is not available in base R, we have written our own. 
- However, we can get functions from other sources, called libraries

- libraries are collection of functions, generaly specialized task ; for example the package `ggplot2`, which we will be heavily using throughout this course is a library that contains functions for makes plots.

- To use a library, load it using the `library` function
```{r}
library(RBedtools)
```
- most libraries are be hosted on CRAN, and can be installed in R using the `install.packages` function

## the Tidyverse

- the tidyverse is a collection of packages that provide *fast, well documented* functions for data science.
- This inlucdes reading and writing data, organizing and "tidying" data, and plotting data
```{r}
library(tidyverse)
```

- tibble: a new data type analgous to a `data.frame`, but with more consistent indexing, and no type coerscion
- dplyr: a series of functions for subsetting, filtering, changing, and summarizing dataframes/tibbles
- tidyr: functions for tidying data(more on that later)
- stringr: library for efficiently working with strings
- readr: efficiently read/write data
- forcats: library for efficiently working with factors
- purrr: adds common programming methods from other languages into







