---
title: "Working with the Tidyverse"
output: html_notebook
---

We're going to practice using some of the functions in the `dplyr` library. First, lets install the `tidyverse` packages
```{r}
install.packages('tidyverse')
```

You will be using the pipe function all lot; I recommend setting a key board short cut for it. Go to Tools > modify keyboard searchcuts > search for "pipe"  >  click to change the command. I find `alt + 2` or `option + 2` work well.

load the `dplyr` library
```{r}

```

we'll be working with the `starwars` dataset

```{r}
data("starwars")
starwars
```

Select the `Name` column from the `starwars` dataset. Store this in a variable 

```{r}

```

now `pull` the `Name` column from the dataset. store this in a variable

```{r}

```

Notice the differences in the returned data. use `class` to examine both outputs 

```{r}

```

select the name and height columns 

```{r}

```

 Using pipes to add to the previous command,  add a column with the height in inches.  the height column is currently in cm and  1 cm is 0.393701 inches 
```{r}

```

plot a scatter plot for the height in inches 
```{r}

```

using the main data set, select just the `skin_color` column. `group_by` this column and calculate the number of observations in each group. (hint use the `n()` function)

```{r}

```

## Tying it all together
Lets practice what we'ved learned on real data

In the main folder for this project, there is a subfolder called `src`; inside this is the file "tx_expression_subset.tsv.gz". This is some data from a real RNA-seq experiment and contains transcript expression values; there are two columns for the gene and transcript name, and the rest of the columns are the samples each transcript is expressed in.
In order, do the following;
 - read the data into R, using the absolute path for the file as the input file
 - reshape all the data except the `gene_name` and `transcript_id` columns into long format
 - calculate per sample total gene expression - for each gene in each sample calculate the sum of all associated transcripts(hint: you will need to group by multiple columns)
 - filter the data so you only have data for the following genes: DCX , GRM3 , MYO7A
 - make a boxplot of expression values for these genes; color each box plot by gene expression

```{r}

```
 










