---
title: "Working with the Tidyverse"
output: html_notebook
---

We're going to practice using some of the functions in the `dplyr` library. First, lets install the `tidyverse` packages
```{r}
install.packages('tidyverse')
```

You will be using the pipe function all lot; I recommend setting a key board short cut for it. Go to Tools > modify keyboard searchcuts > search for "pipe"  >  click to change the command. I find `alt + 2` or `option + 2` work well.

load the `dplyr` library
```{r}

```

in the `starwars` dataset, keep only values about Humans from the homewrold Tatooine. Store the result in variable 

```{r}

```

select only the height `Name` and height columns from the previous result

```{r}

```

repeat the first two excerises, but chain them together using pipes (` %>% `)

```{r}

```

keep adding to the chained commands, and convert the Height column to inches. the height column is currently in cm and  1 cm is 0.393701 inches 
```{r}
dir('../src')
```



reshape the `iris` data frame in to long format with `pivot_longer` ; calculate the max an min for each column using `group_by` and `summarise`

## Tying it all together
Lets practice what we'ved learned on real data;

In the main folder for this project, there is a subfolder called `src`; inside this is the file "tx_expression_subset.tsv.gz". This is some data from a real RNA-seq experiment and contains transcript expression values; there are two columns for the gene and transcript name, and the rest of the columns are the samples each transcript is expressed in.
In order, do the following;
 - read the data into R, using the absolute path for the file as the input file
 - reshape all the data except the `gene_name` and `transcript_id` columns into long format
 - calculate per sample total gene expression - for each gene in each sample calculate the sum of all associated transcripts(hint: you will need to group by multiple columns)
 - filter the data so you only have data for the following genes: DCX , GRM3 , MYO7A
 - make a boxplot of expression values for these genes; color each box plot by gene expression

```{r}

```
 










