---
title: "R Notebook"
output: html_notebook
---

## install complexHeatmap
```{r}
install.packages(c('BiocManager','viridisLite' ) )
BiocManager::install('ComplexHeatmap')
library(ComplexHeatmap)
library(tidyverse)
```



- here is some example data we will be using
- log transforming gene expression data is commonly done before plotting.
```{r}
gene_expression_matrix <- readRDS('../src/gene_expression_ss.RDS') %>% {log2(.[1:24, 1:24]+1)}
```

## A single heatmap 

- plot a heat map of the data 
```{r}
Heatmap(gene_expression_matrix)
```

- set the colors of the heat map using `colorRamp2`
```{r}

library(circlize)
col_fun = colorRamp2(c(0, 5, 15), c("purple", "green", "yellow"))
col_fun(seq(0, 15))
Heatmap(gene_expression_matrix, name = "mat", col = col_fun)
class(col_fun)

```

- set the color to the viridis palette. you can generate a vector of viridis colors by using `viridisLite::viridis`

```{r}

library(viridisLite)
viridis_pal =viridisLite::viridis(100)
Heatmap(gene_expression_matrix, name = "mat", col = viridis_pal)

```

- change the heatmap such that there is a white gap between each tile.
```{r}
Heatmap(gene_expression_matrix, col = viridis_pal,
        rect_gp = gpar(col = "white", lwd = 2))
?Heatmap
```

- create row and column titles accordingly 

```{r}

```


- only cluster the columns

```{r}

```

## 2.7 Heatmap split
- split the heatmap into 3 subpieces, by column


## Heatmap annotations 
- this file contains metadata about the samples in gene expression data. sample_accession corresponds to the name of each sample in the gene expression matrix.

```{r}
sample_metadata <- read_tsv('../src/gene_expression_sample_tsv.gz')
```

- create a simple column annotation that corresponds to the `Tissue` column in the sample metadata

```{r}

```



