---
title: "Practice with maps "
output: html_notebook
---



Lets make a plot with some real data. We're gonna be working with some data about COVID-19 testing in maryland. 
'day_2/MD_COVID_cases_by_zip.tsv.gz' contains the per day number of positive covid cases by zip code for about 160 days in the state of maryland. 
'day_2/maryland_zipcode_to_county.tsv.gz' contains a mapping for maryland zip code to counties
'day_2/maryland_counties.shp' is a shapefile for counties in maryland

Were going to take this data and make this:

```{r}

```

As a starting point, this is the generaly workflow. Load data > join county level data to zip level data > aggreagte data to the county level  >  join map data > make animation 



- first add the county information to the covid data. check how many counties are shared between the covid data and the map data 
```{r}

```


- aggregate zip level covid data to county level
```{r}

```

- the character date to a DateTime vector
```{r}

```


- add the map data to the covid data. filter out the date '2020-05-21' and plot it 
```{r}

```



