---
title: '**BIOF 076: Visualization with R**'
author: " Instructor: Vinay Swamy"
date: "October 5th-9th"
output: 
  word_document:
    default
  pdf_document:
    default
subtitle: Creating publication quality figures and interactive web apps with the R
  programming language
---

# Introduction
Creating compelling visualizations is an important aspect of biomedical research. The R programming language provides many libraries for creating beautiful figures and interactive web apps. As R is an open source project, it facilitates open science and reproducible research. R has been heavily used by bioinformaticians and data scientists for years, and has become increasingly easy to use. This course is designed to allow researchers to quickly dive into R and make visualizations for their own work. No previous experience in R is required. 

After this course you will be able to:

* Load and prepare data for plotting
* Generate common scientific plots like Bar graphs, scatter plots, and heat maps using multiple plotting libraries 
* Use git, GitHub and binder to share plots
* Use visualizations to explore new data
* Combine multiple plots to create publication quality figures 
* Design interactive web apps with R-Shiny 
* Integrate plots from R into posters and papers
* Complete a final project with your own data

# Format
The workshop is structured as a series of interactive lessons, with a lecture and exercises components. There 
Our engagement during this workshop will take several forms: 

* Class materials: All materials, including lecture slides and excercises will be availble on canvas, and the course Github repository
* All lessons will be held live over Zoom
* Communications: There will be a slack group created for the class.
 
# Software and Materials 
We will have two sessions to install all software before the course at **10am** and **4pm** on **Friday, October 2nd**. Some of the main software we will be using:

* Software
  + R language base system -  the core interpreter for the R language that runs the code we will write
  + Rstudio - an integrated devleopment environment(IDE) that makes it signficantly easier to write code
  + git - a version control system for writing code
  + GitHub - students will sign up for GitHub, an online repository for code. 
  + GDAL - software for using maps in R
* Materials
  + A computer, ideally with administrative access. (you likely *do not* have administrative access on a government issued computer)
  + Multiple screens(2 monitors, computer  +  tablet/phone etc)

# Schedule

```{r, echo=FALSE, include=F}
library(tidyverse)
library(knitr)
library(kableExtra)
sch <- 'Monday	9AM-9:50AM	Course Introduction/ Basic Programming in R - Part 1
Monday	10AM-10:50AM	Basic Programming in R - Part 2
Monday	11AM-12PM	Using the ggplot2 library - Basic Plots 
Monday	12PM-1:00PM	Break
Monday	1PM-1:50PM 	Manipulating Data with the tidyverse Part 1
Monday	2PM-2:50PM	Using the ggplot2 library - Customizing themes and Aesthetics
Monday	3pm-3:50PM	Office hours 
Tuesday	9AM-9:50AM	Review / Conditional programming 
Tuesday	10AM-10:50AM	Manipulating Data with the tidyverse Part 2
Tuesday	11AM-12PM	Using the ggplot2 library - Complex Plots 
Tuesday	12PM-1:00PM	Break 
Tuesday	1PM-1:50PM 	Extensions to Ggplot 
Tuesday	2PM-2:50PM	Plotting with Maps  and making Animated plots 
Tuesday	3pm-3:50PM	Office hours
Wednesday	9AM-9:50AM	Review/ Making Heatmaps 
Wednesday	10AM-10:50AM	Combining multiple plots / Intro to R Markdown 
Wednesday	11AM-12PM	Designing plots to accurately represent data
Wednesday	12PM-1:00PM	Break 
Wednesday	1PM-1:50PM 	Making Copycat Plots  - Building intuition for making novel plots 
Wednesday	2PM-2:50PM	Interactive plots with Plotly
Wednesday	3pm-3:50PM	Office Hours 
Thursday	9AM-9:50AM	Review/Intro To Shiny 
Thursday	10AM-10:50AM	Shiny - UI
Thursday	11AM-12PM	Shiny - server
Thursday	12PM-1:00PM	Break 
Thursday	1PM-1:50PM 	Deploying Shiny apps
Thursday	2PM-2:50PM	Complex Shiny Apps 
Thursday	3pm-3:50PM	Office Hours
Friday	9AM-9:50AM	Course Summary 
Friday	10AM-10:50AM	Student Project  Development
Friday	11AM-12PM	Student Project  Development
Friday	12PM-1:00PM	Break 
Friday	1PM-1:50PM 	Student project presentations 
Friday	2PM-2:50PM	Student project presentations 
Friday	3pm-3:50PM	Office Hours'
schedule <- read.table(text = sch,sep = '\t', col.names = c('Day','Time','Topic'))
```
```{r, echo=F}
kbl(schedule, booktabs = T, longtable=T) %>% 
        collapse_rows(1,latex_hline = "major", valign = "middle")
```

# Office Hours
Office hours will be held at the end of each day from 3-4PM. 

# FAQ

Q. Do I need any Prior Experience in R

A. No, This class requires NO expreience in R. We will cover everything you need to know within the course.

Q. I don't have administrative access to my computer, how will I be able to install the necessary software?

A. While it's best to work on your own machine, a standalone cloud based environment will be available for people to use.





