---
title: "extending ggplot"
output: html_notebook
---


## ggpattern

```{r}
library(gapminder)
library(tidyverse)
gapminder_bygender <- gapminder %>% 
    mutate(gender = sample(c('male', 'female'), nrow(gapminder), replace = T))
```

I've added some data about gender to the `gapminder` dataset. use this data set to calculate the average lifespan per continent per gender. plot a bargraph with colors for each continents and patterns for each gender. 

```{r}
remotes::install_github("coolbutuseless/ggpattern")
```


## revisting the line graph from earlier 
use `ggline` to remake the line graph we made a few exercises ago

```{r}

```

filter the data to the last time point, and make a boxplot with `ggboxplot`. Include a test for a statstically significant change in the data on the plot.  

##  external themes and palettes.
- there are many libraries that add new themes to `ggplot`, too many for us to cover. 
- two packages we will cover are `ggthemes` and `ggsci`



## gganimate


## plotting with maps 

Lets make a plot with some real data. We're gonna be working with some data about COVID-19 testing in maryland. 
'day_2/MD_COVID_cases_by_zip.tsv.gz' contains the per day number of positive covid cases by zip code for about 160 days in the state of maryland. 
'day_2/maryland_zipcode_to_county.tsv.gz' contains a mapping for maryland zip code to counties
'day_2/maryland_counties.shp' is a shapefile for counties in maryland

Were going to take this data and make this:

```{r}

```

As a starting point, this is the generaly workflow. Load data > join county level data to zip level data > aggreagte data to the county level  >  join map data > make animation 



- first add the county information to the covid data. check how many counties are shared between the covid data and the map data 
```{r}

```


- aggregate zip level covid data to county level
```{r}

```

- the character date to a DateTime vector
```{r}

```


- add the map data to the covid data. filter out the date '2020-05-21' and plot it 
```{r}

```

