---
title: "Interactive Plots"
author: "Vinay Swamy"
date: "9/26/2020"
output:
  ioslides_presentation: 
    widescreen: true
---

<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

## Interactive plots are tricky
- An interactive plot is any plot that is reactive to use input;
- It can be diffcult to both develop and share interactive plots, as extensive libraries are needed for both of these parts
- There are a couple of packages for making interactive plots in R. While these packages make it relatively easy to make interactive plots, sharing them cna be a little more tricky, and We'll cover it a little more in future classes

## `ggiraph`
- A simple and lightweight interactive plotting library is `ggiraph`. It use `ggplot` style syntax, and provides some drop in replacement for some `ggplot` geoms 

```{r, include=FALSE}
library(tidyverse)

```

```{r}
library(ggplot2)
library(ggiraph)
data <- mtcars
data$carname <- row.names(data)
gg_point = ggplot(data = data) +
    geom_point_interactive(aes(x = wt, y = qsec, color = disp,
    tooltip = carname, data_id = carname)) + 
     theme_minimal()
girafe(ggobj = gg_point)
```

- `ggiraph` is limited to adding tooltips ( info displayed on hover), and can't really do much beyond that 

## Plotly 
- Plotly is a very powerful library for making interactive plots. Its a little weird in that plotly is developed by a independent company, but makes their library open source. If you use plotly, they do add a small watermark to your plots

-Most of plotly is actually a javascript library, and the functions in R  are wrapper on top of the java script code.

- but its worth it, because you can create some pretty cool plots 
```{r}
library(plotly)

mtcars$am[which(mtcars$am == 0)] <- 'Automatic'
mtcars$am[which(mtcars$am == 1)] <- 'Manual'
mtcars$am <- as.factor(mtcars$am)

fig <- plot_ly(mtcars, x = ~wt, y = ~hp, z = ~qsec, color = ~am, colors = c('#BF382A', '#0C4B8E'))
fig <- fig %>% add_markers()
fig <- fig %>% layout(scene = list(xaxis = list(title = 'Weight'),
                     yaxis = list(title = 'Gross horsepower'),
                     zaxis = list(title = '1/4 mile time')))

fig
```

- plotly has a ggplot-esque API (thank you [yigit Erol](https://medium.com/swlh/step-by-step-data-visualization-guideline-with-plotly-in-r-fbd212640de2) for the example)

```{r}
library(plotly)
x <- c(2015:2020)
Beverages <- rnorm(x, mean = 10)
Grocery <- rnorm(x, mean = 1)
Snack <- rnorm(x, mean = 5)
x <- c(2015:2020)
df <- data.frame(x, Beverages, Grocery, Snack)
df
```

- now the actual plot. `plot_ly` is very loose with how it populates data, so what ever parameters we don't specify, it will try and guess from the data we provide 

```{r}
fig <- plot_ly(data = df, x = ~x, marker=list(size=10))
fig
```

- we didn't provide a mapping what type of plot we wanted, and so it defaults to a bar graph

```{r}
 plot_ly(data = df, x = ~x,type = 'scatter', marker=list(size=10))
```
-we told it scatter plot, but didn't provide any Y data


```{r, warning=F, message=F}
plot_ly(data = df, x = ~x,tmarker=list(size=10)) %>% 
    add_trace(y = ~Beverages, name = 'Beverage',mode = 'lines') %>% 
    add_trace(y = ~Grocery, name = 'Grocery', mode = 'lines+markers') %>% 
    add_trace(y = ~Snack, name = 'Snack', mode = 'markers') %>% 
    layout(title = 'Sales by Catergories Between 2015 & 2020', 
                      yaxis = list(title = 'Average Sales'),
                      xaxis = list(title = 'Time') ) 
```
-  note that we build graphs iteratively, unlike `ggplot` where all plots of a certain type are specified at once 

- plotly does provide function to auto convert `ggplots` into `plotly` plots 
```{r}
df_long <- df %>% pivot_longer(-x)
gg <- ggplot(df_long) + 
    geom_line(aes(x=x, y=value, color = name))
gg
```
```{r}
gg %>% ggplotly()
```
- This only works for simple plots; the 3D plot above couldn't be created like this, because `ggplot` only understands `x` and `y` aesthetics

- Between these two libraries, you should have enough to start making your own plots




