---
title: '**BIOF 076: Visualization with R**'
author: " Instructor: Vinay Swamy"
date: "January 10th-14th"
output: 
  pdf_document:
    default
subtitle: Creating publication quality figures and interactive web apps with the R
  programming language
---

# Introduction
Creating compelling visualizations is an important aspect of Biomedical research. The R programming language provides many libraries for creating beautiful figures and interactive web apps. As R is an open source project, it facilitates open science and reproducible research. R has been heavily used by bioinformaticians and data scientists for years, and has become increasingly easy to use. This course is designed to allow researchers to quickly dive into R and make visualizations for their own work. No previous experience in R is required. 

After this course you will be able to:

* Load and prepare data for plotting
* Generate common scientific plots like Bar graphs, scatter plots, and heat maps using multiple plotting libraries 
* Use git, GitHub and binder to share plots
* Use visualizations to explore new data
* Combine multiple plots to create publication quality figures 
* Design interactive web apps with R-Shiny 
* Integrate plots from R into posters and papers
* Complete a final project with your own data

# Format
The workshop is structured as a series of interactive lessons, with a lecture and exercises components. There 
Our engagement during this workshop will take several forms: 

* Class materials: All materials, including lecture slides and excercises will be availble on canvas, and the course Github repository
* All lessons will be held live over Zoom
* Communications: There will be a slack group created for the class.
 
# Software and Materials 
We will install all software at the beginning of the course. This will require administrative access on your computer. If you do not have this(ie, you have a government machine), the software you will need is listed below. 

* Software
  + R language base system -  the core interpreter for the R language that runs the code we will write
  + Rstudio - an integrated development environment(IDE) that makes it significantly easier to write code
  + git - a version control system for writing code
  + GitHub - students will sign up for GitHub, an online repository for code. 
  
# Schedule

```{r, echo=FALSE, include=F}
library(tidyverse)
library(knitr)
library(kableExtra)
sch <- 'Monday,9AM-9:50AM,Course Introduction/Basic Programming in R - Part 1
Monday,10AM-10:50AM,Basic Programming in R - Part 2
Monday,11AM-12PM,Basic Programming in R - Part 3
Monday,12PM-1:00PM,Break
Monday,1PM-1:50PM,Using the ggplot2 library - Basic Plots 
Monday,2PM-2:50PM,Using the ggplot2 library - Customizing themes and Aesthetics
Monday,3pm-3:50PM,Office hours 
Tuesday,9AM-9:50AM,Review / Conditional programming 
Tuesday,10AM-10:50AM,Manipulating Data with the tidyverse Part 1
Tuesday,11AM-12PM,Manipulating Data with the tidyverse Part 2
Tuesday,12PM-1:00PM,Break 
Tuesday,1PM-1:50PM,Manipulating Data with the tidyverse Part 3  
Tuesday,2PM-2:50PM,Using the ggplot2 library - Complex Plots
Tuesday,3pm-3:50PM,Office hours
Wednesday,9AM-9:50AM,Extensions to ggplot 
Wednesday,10AM-10:50AM,Combining multiple plots to make Figures 
Wednesday,11AM-12PM,Plotting with Maps and making Animated plots 
Wednesday,12PM-1:00PM,Break 
Wednesday,1PM-1:50PM,Intro to R Markdown 
Wednesday,2PM-2:50PM,Making Copycat Plots - Building intuition for making novel plots 
Wednesday,3pm-3:50PM,Office Hours 
Thursday,9AM-9:50AM,Review/Interactive plots with Plotly
Thursday,10AM-10:50AM,Intro To Shiny 
Thursday,11AM-12PM,Shiny - UI
Thursday,12PM-1:00PM,Break 
Thursday,1PM-1:50PM,Shiny - server
Thursday,2PM-2:50PM,Complex Shiny Apps 
Thursday,3pm-3:50PM,Office Hours
Friday,9AM-9:50AM,Course Summary 
Friday,10AM-10:50AM,Student Project  Development
Friday,11AM-12PM,Student Project  Development
Friday,12PM-1:00PM,Break 
Friday,1PM-1:50PM,Student project presentations 
Friday,2PM-2:50PM,Student project presentations 
Friday,3pm-3:50PM,Office Hours'

schedule <- read.table(text = sch,sep = ',', col.names = c('Day','Time','Topic'))

```
```{r, echo=F}
kbl(schedule, booktabs = T, longtable=T) %>% 
        collapse_rows(1,latex_hline = "major", valign = "middle")
```

#### A note on the schedule 
- We will try to cover all material listed in the schedule. However, it is very possible we will move slower than anticipated, and so any material we do not cover within the first 4 days will be covered on the final day(Friday) in lieu of the presentations

# Office Hours
Office hours will be held at the end of each day from 3-4PM. 

# FAQ

Q. Do I need any Prior Experience in R

A. No, This class requires NO expreience in R. We will cover everything you need to know within the course.

Q. I don't have administrative access to my computer, how will I be able to install the necessary software?

A. While it's best to work on your own machine, a standalone cloud based environment will be available for people to use.





