---
title: "Working with the Tidyverse"
output: html_notebook
---

We're going to practice using some of the functions in the `dplyr` library. First, lets load the `tidyverse` packages

```{r  ex_1}

```

You will be using the pipe function all lot; I recommend setting a key board short cut for it. Go to Tools > modify keyboard search cuts > search for "pipe"  >  click to change the command. I find `alt + 2` or `option + 2` work well.

load the `dplyr` library

- `select` select columns from a dataframe
- `filter` subset a dataframe based on a logical operator or vector
- `mutate` to add or change columns in a data frame
- `arrange` to sort a dataframe based on a column
- `pull` extract a column from a dataframe as a vector
- `rename` rename a column 

we'll be working with the `starwars` dataset. Run this chunk to load the data 

```{r}
data("starwars")
starwars
```



```{r ex_2}

```

Select the `name` column from the `starwars` dataset. Store this in a variable 

```{r ex_3}

```

now `pull` the `name` column from the dataset. store this in a variable

```{r ex_4}

```

Notice the differences in the returned data. use the `class` function to examine both outputs 

```{r ex_5}

```
select the name and height columns 
```{r ex_6}


```

 Using pipes to add to the previous command,  add a column with the height in inches.  the height column is currently in cm and  1 cm is 0.393701 inches 
```{r ex_7}

```

adding to the same pipe command, rename the height colunmn to height_cm

```{r ex_8}

```


filter the starwars dataframe to get every person who has blue eyes 

```{r ex_9}

```

filter the starwars dataframe to get rows whose homeworld is either Tatooine or Naboo	
hint: use %in%

```{r ex_10}


```

use groupby and summarise to calculate the avergae height for each eye_color, sex group

```{r ex_11}

```




## Example - GTF file

- a GTF file contains information about the location of genes within the genome, as well as vartiety of other annotations. `../src/gtf_example.tsv.gz` contains 5 genes. 
- Each gene has multiple transcripts, and each transcript has a protein coding region, or CDS.  read this data in, and use functions from the tidyverse to calculate the length of each CDS, and determine the gene with the longest CDS. 
steps:
    - read the data (remember the reading functions from readr)
    - filter values that only correspond the to CDS(look at the `type` column) 
    - group_by `transcript_id` and calculate the smallest start and the largest end(recall that `min` returns the 
    smallest value in a vector, and `max` the largest)
    - add a column for the CDS length (the largest end - smallest start  = length)

```{r ex_12}


```


## reshaping data

practice going from wide to long format using the pivot_longer function, using the economics data frame


```{r}
data("economics")
```

## Tying it all together

- the file  "../src/tx_expression_subset.tsv.gz" has some data from a real RNA-seq experiment and contains transcript expression values; there is a column for transcript name, and the rest of the columns are the samples each transcript is expressed in. 

- pivot the data from wide to long

- Use the gtf data from above to make a mapping between transcript_id and gene_name, ensuring that we have distinct transcript_id's. Hint( use `select` and `distinct`)

- filter this data down to these 3 genes: DCX , GRM3 , MYO7A

- join this gene<>tx mapping to the RNA-seq data.

- group the data by gene_name and sample to calculate the total expression per gene,per sample

- add a column for log transformed gene expression ( formula:  log_transformed_data = log2(data +1))

- make a box plot for the log scaled gene expression, coloring by gene 


```{r ex_13}




```
 




