---
title: "Intro to Shiny"
author: "Vinay Swamy"
date: "9/28/2020"
output:
  ioslides_presentation: 
    widescreen: true
---

<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>
## R-Shiny
- R-Shiny is a series of packages for creating interactive, web-based apps, which can run locally or as a stand alone website.

## what can you do with Shiny
- [EyeIntegration](https://eyeintegration.nei.nih.gov)

## what can you do with Shiny
- Before we get to shiny, lets talk a little bit about interactivity and websites

## Interactive apps
- In general, designing an interactive app is fairly difficult
- Interactivity requires a lot of code, so distributing a standalone app like Microsoft Word, or Rstudio requires a lot of overhead.

## Interactivity on the web
- Websites by their very nature require a lot of interactivity, and so code for handling it has been well developed and established, in languages like `CSS` and `JavaScript` 
- In the early days of interactivity in R, people found it much easier to use the existing code from web development, rather than re-write everything themselves. 
- this is why writing shiny code has a lot of parallels in web development

## How does a website work 
- Understanding how a website works will help a lot with writing shiny apps. 
- Every website has two main components - a server(backend) and a User Interface(UI/frontend).
- The UI handles all the interacttivity, scrolling, text input, buttons, etc.
- The server is where data required to run the app is stored, and where computation to power the site is run.

## Example - Google
- [Google](https://www.google.com)


## Server/ UI summary
- the UI waits for us to do something (input),
- this input is passed to the server, which then performs actions based on the input, and populates an output within the UI

- This happens in a loop extremely quickly


## Whats a simple Shiny App look like?

## Whats a simple Shiny App look like?
- `shinyApp(ui = ui, server = server)` is what actually runs the server and UI

## the UI 
- the UI that we write is a function,  where each part of the UI is an argument
```{r eval = F}
library(shiny)
ui <- fluidPage(
    # App title ----
    titlePanel("Hello Shiny!"),
    # Sidebar layout with input and output definitions ----
    sidebarLayout(
        # Sidebar panel for inputs ----
        sidebarPanel(
            # Input: Slider for the number of bins ----
            sliderInput(inputId = "bins", label = "Number of bins:", min = 1,max = 50, value = 30)
        ),
        # Main panel for displaying outputs ----
        mainPanel(
            # Output: Histogram ----
            plotOutput(outputId = "distPlot")
        )
    )
)
```

## the UI 
- elements inside the UI can composed of smaller elements; in this case, `sidebarLayout` is composed of a `sidebarPanel` and a `mainPanel`
- be **VERY** careful with parentheses and commas inside the UI function. This is probably the most common error when writing a shiny app. Using this indent style , and always putting commas directly after parentheses - `),` vs `)  ,` - help mitigate this.

## UI elements
- All *interactive* and *reactive* elements within the UI have an ID associated with it. 
- Inputs are generally interactive, and outputs are reactive
```{r eval = F}
sliderInput(inputId = "bins", label = "Number of bins:", min = 1,max = 50, value = 30)
plotOutput(outputId = "distPlot")
```

## UI elements 
-  we can then access each element by its id in in the sever, and similarly set each output

## the server
- unlike the UI, which is itself a predefined function, the server is a unique function that we design for each app 
```{r eval = F}
server <- function(input, output) {
    output$distPlot <- renderPlot({ 
        x    <- faithful$waiting
        bins <- seq(min(x), max(x), length.out = input$bins + 1)
        
        hist(x, breaks = bins, col = "#75AADB", border = "white",
             xlab = "Waiting time to next eruption (in mins)",
             main = "Histogram of waiting times")
        
    })
    
}
```
- the `input` and `ouptut` *must* be specified as arguments, and are almost always the only arguments for the function.

## the server input
- `input` is a list where each element corresponds to an `inputID` within the UI. The value within each inputID corresponds to the specific UI element its associated with 
```{r, eval = F}
## from UI
sliderInput(inputId = "bins", label = "Number of bins:", min = 1,max = 50, value = 30)
## from server 
bins <- seq(min(x), max(x), length.out = input$bins + 1)
```

## the server output
- the `output` argument contains "empty" outputIDs, that are connected to the outputs in UI
- outputs must alway be some kind `render*` object
```{r, eval = F}
output$distPlot <- renderPlot({ 
        x    <- faithful$waiting
        bins <- seq(min(x), max(x), length.out = input$bins + 1)
        
        hist(x, breaks = bins, col = "#75AADB", border = "white",
             xlab = "Waiting time to next eruption (in mins)",
             main = "Histogram of waiting times")
        
    })
```

## server formats
```{r eval=F}
server <- function(input, output) {
    output$distPlot <- renderPlot({ 
        x    <- faithful$waiting
        bins <- seq(min(x), max(x), length.out = input$bins + 1)
        
        hist(x, breaks = bins, col = "#75AADB", border = "white",
             xlab = "Waiting time to next eruption (in mins)",
             main = "Histogram of waiting times")
        
    })
    
}
```
- notice that server relies a lot more on expressions, chunks of code demarcated by `{}`. This makes it much easier to write server code. 
- Remember that any code invloving computations must be run in server; the UI is only for UI, and server is for everything else

## writing shiny code
- Shiny code works best in an `Rscript`. This is nothing more than a text file with a `.R` extension. We won't be able to run code interactively like a notebook
- However Rstudio provides a little help to run shiny apps. 
- Open the file 'shiny_examples/basic_app.R'

## loading data when using shiny

- make sure you read in all the data you need at the beginning of your script. If you read data in within shiny code, the data will be run in constanly and get very slow.

## more control over layout - columns
- as opposed to a single column with `mainPanel`, you can designate multiple columns within `fluidPage`
with the `column` function
```
fluidPage(
    
    column(3,
           radioButtons("radio", "Radio buttons",
                        choices = list("Choice 1" = 1, "Choice 2" = 2,
                                       "Choice 3" = 3),selected = 1),
           plotOutput(outputId = "distPlot")),
                        
                                      
    
    column(3,
           selectInput("select", "Select box", 
                       choices = list("Choice 1" = 1, "Choice 2" = 2,
                                      "Choice 3" = 3), selected = 1),
           plotOutput(outputId = "distPlot"))
        )
```
- the first number in the column corresponds to the width; the width of one page is 12, so these columns each take up a quarter of the screen



