---
title: "R Notebook"
output: html_notebook
---

## install complexHeatmap
```{r}
install.packages(c('BiocManager','viridisLite' ) )
BiocManager::install('ComplexHeatmap')
library(ComplexHeatmap)
library(tidyverse)
```

- here is some example data we will be using
- log transforming gene expression data is commonly done before plotting.
```{r}
gene_expression_matrix <- readRDS('../src/gene_expression_ss.RDS') %>% {log2(.[1:24, 1:24]+1)}
```

## A single heatmap 

- plot a heat map of the data 
```{r}
Heatmap(gene_expression_matrix)
```

- set the colors of the heat map using `colorRamp2`
```{r}
library(circlize)

col_fun = colorRamp2(c(0, 2, 4), c("purple", "green", "yellow"))
col_fun(seq(-3, 3))
Heatmap(gene_expression_matrix, col = col_fun)

```

- set the color to the viridis palette. you can generate a vector of viridis colors by using `viridisLite::viridis`

```{r}
library(viridisLite)
viridisLite::viridis
colors <- viridisLite::viridis(25)

Heatmap(gene_expression_matrix, col = colors)
```

- change the heatmap such that there is a white gap between each tile.
```{r}
Heatmap(gene_expression_matrix, col = colors, rect_gp = gpar(col = "white", lwd = 2))
```

- create row and column titles accordingly 

```{r}
Heatmap(gene_expression_matrix, col = colors, rect_gp = gpar(col = "white", lwd = 2), row_title = 'Gene_name', column_title = 'Samples')
```


- only cluster the columns

```{r}
Heatmap(gene_expression_matrix, col = colors, rect_gp = gpar(col = "white", lwd = 2), row_title = 'Gene_name', column_title = 'Samples', cluster_rows = F, cluster_columns = T)
```

## 2.7 Heatmap split
- split the heatmap into 3 subpieces, by column

```{r}
column_km = 3)
Heatmap(gene_expression_matrix, col = colors,column_km = 3,  row_title = 'Gene_name', column_title = 'Samples', cluster_rows = F, cluster_columns = T)
```


## Heatmap annotations 
- this file contains metadata about the samples in gene expression data. sample_accession corresponds to the name of each sample in the gene expression matrix.

```{r}
sample_metadata <- read_tsv('../src/gene_expression_sample_tsv.gz')
```

- create a simple column annotation that corresponds to the `Tissue` column in the sample metadata

```{r}
ha <- HeatmapAnnotation(Tissue= sample_metadata$Tissue)
Heatmap(gene_expression_matrix, col = colors,column_km = 3,  row_title = 'Gene_name', column_title = 'Samples', cluster_rows = F, cluster_columns = T, bottom_annotation = ha)
```



