---
title: "Customizing GGplots "
author: "Vinay Swamy"
date: "9/28/2020"
output:
  ioslides_presentation: 
    widescreen: true
---

<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.height = 5, fig.width =5)
options(warn = -1)
```
```{r, include=F}
library(patchwork)
library(RColorBrewer)
library(pals)
library(tidyverse)
plot_pal <- function(cols){
  pdata <- tibble(col = cols, x= seq(1, length(cols)), y=0)
  cv <- cols
  names(cv) <- cv
  plot <- ggplot(pdata) + 
    geom_tile(aes(x=x, y=y,fill = col), width = 1, height= .5) +
    scale_fill_manual(values=cv) +
    theme_void()+
    theme(legend.position = 'none')
  return(plot)
}

```

## setting graphical parameters
- up until now we have been using the default graphical parameters.
- graphical parameters include axis names, legend size/orientation, plot titles, grid lines etc
- all graphical parameters can be set with the `theme` function. 
- lets take a look at the `theme` [documentation](https://ggplot2.tidyverse.org/reference/theme.html)
- all the `theme_...` functions we've been using are preset configurations of a theme

## Simple Theme modifiers
- There are several functions that allow you to modify specific elements, designed for simplicity
- `xlab`/`ylab` - change the axis labels 
- `ggtitle` -  add a title to a plot
- `xlim`/`ylim` - change axis limits

## Simple Theme modifiers

```{r}
data(mpg)
mpg
ggplot(mpg) + # standard plot
  geom_bar(aes(x=manufacturer))
```

## Simple Theme modifiers
- change axis labels , and add a title 
```{r}
data(mpg)
mpg
```

## Simple Theme modifiers

```{r}
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer)) + 
  xlab( 'Car Manufacturer' ) + 
  ylab( 'Number of Cars' ) + 
  ggtitle('Cars built in 2018 Q1 by Manufacturer')+
  theme_minimal()
```

## changing text with `theme`

- we can see in the `theme` function there are multiple inputs for textual parts of the text like `plot.title`, or `axis.text.x`
- there are various setting for text - font, size color etc; These are controlled by the `element_text` function, the output of which we assign to to specific graphical parameter, like `axis.text`. these parameters change everything about the text, except the text itself; setting the text is done with functions like `xlab` or `ggtitle` or populated directly from the data

## changing text with `theme`
- the previous plot had squished text on the x axis 
```{r}
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer)) + 
  xlab('Car Manufacturer') + 
  ylab('Number of Cars' ) + 
  ggtitle('Cars built in 2018 Q1 by Manufacturer') +
  theme(axis.text.x = element_text(size = 12, angle = 45))
```

- Note that background of the plot also changed; this is because we were previously using `theme_minimal` which has the background set to white, but in `theme` the default is grey; this can be fixed by combined the two themes

## changing text with `theme`

```{r}
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer)) + 
  xlab('Car Manufacturer') + 
  ylab('Number of Cars' ) + 
  ggtitle('Cars built in 2018 Q1 by Manufacturer') +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 12, angle = 45))

```

- the order is important; as each theme is applied in the order it's added, so adding theme_minimal last can sometimes override changes in the theme made previously

## changing text with `theme`
- while a little better, it would be better if the axis text was off the graph
```{r}
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer)) + 
  xlab('Car Manufacturer') + 
  ylab('Number of Cars' ) + 
  ggtitle('Cars built in 2018 Q1 by Manufacturer') +
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1))
```

## setting inheritance
- graphical settings are all linked in a hierarchical manner; for example this is the hierarchy for axis.text.x: `text` > `axis.text` > `axis.text.x` . 
- Changing settings for a one parameter changes everything below it in the hierarchy; however, if you specify changes for parameter below it will take precedence over anything set previously

```{r}
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer)) + 
  xlab('Car Manufacturer') + 
  ylab('Number of Cars' ) + 
  ggtitle('Cars built in 2018 Q1 by Manufacturer') +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), text = element_text(family ='Palatino' ))
```

## guides
- guides are an alternative  method to setting graphical parameters in theme. 
- They were introduced to ggplot after themes, and are generally a more intuitive interface for setting graphical parameters, but lack the full range of parameters offered by `theme`

## guides
- For example, what if we want to change the name of the legend in the graph below

```{r}
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer, fill = class)) + 
  xlab('Car Manufacturer') + 
  ylab('Number of Cars' ) + 
  ggtitle('Cars built in 2018 Q1 by Manufacturer') +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), text = element_text(family ='Palatino' ))

```

## guides
- we have two options: 
- option 1. change the name of the `class` column, or make a new column that has the name we want 

```{r}
mpg$Type_Of_Car <- mpg$class
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer, fill = Type_Of_Car)) + 
  xlab('Car Manufacturer') + 
  ylab('Number of Cars' ) + 
  ggtitle('Cars built in 2018 Q1 by Manufacturer') +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), text = element_text(family ='Palatino' ))
```

## guides
- or option 2, use the `guides` function. For guides we set a value for the aesthetic within the `geom` we want to change ,  in this case `fill`. 
- Next we choose a guide for the graphical parameter associated with that aesthetic we want to change, IE `guide_legend`

```{r}
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer, fill = class)) + 
  xlab('Car Manufacturer') + 
  ylab('Number of Cars' ) + 
  ggtitle('Cars built in 2018 Q1 by Manufacturer') +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1), text = element_text(family ='Palatino' )) + 
  guides(fill = guide_legend(title = 'Type of Car', title.position = 'left'))
```

- while the text of legend cannot be changed inside theme, the position can be 

## colors in R
- color palettes refer the to the set of colors you use within your plot.
- choosing the right color scheme for plot is an under appreciated, but crucial step.

```{r}
length(colors())
```

- there are 657 distinct colors (though 200 of these are different shades of grey)

## What considerations should go into picking a color
- There's been a lot research on what types of colors are best for plots
-  TLDR - plots using color work best when there are *divergent* colors. 
- This can be especially difficult when choosing a continuous fill, because certain colors don't work well as a gradient
- Consider the following palette

## blues

```{r}
plot_pal(cols =pals::brewer.blues(12))
```

## two color gradients 

```{r}
plot_pal(pals::brewer.rdbu(12) )
```

## three color gradients - `viridis`

```{r}
plot_pal(pals::viridis(12))
```

## three color gradients - `magma`

```{r}
plot_pal(pals::magma(12))
```

## three color gradients
- Use three color scales work best for almost all color gradients, and are the most distinguishable 
- these two specific scales in particular, `viridis` and `magma`

## four color gradients

```{r}
plot_pal(pals::jet(30))
```

- the four color gradient looks cool and is tempting use, but its a little too busy.

## useful packages to generate color palettes

- the `RColorBrewer` package has arranged some of these into useful palettes

```{r, fig.height=8, fig.width=8}
library(RColorBrewer)

RColorBrewer::display.brewer.all()
```


the `pals` package
- the `pals` package contains the most extensive set of color palettes and contains over a hundred palettes
-the `Alphabet palette` the largest set distinct colors for discrete data 

```{r}
library(pals)
print(plot_pal(alphabet()))
```


## using palettes functions 
- all functions to generate a color palette function roughly the same way. Select a palette, and provide a number of colors to generate. A list of colors, specified in hex format is returned.
- when using the `pals` package, each palette is its own function and requires only the number of colors

```{r}
alphabet(4)
```


## using color palettes within ggplot 
- ggplot provides a couple methods to change the coloring of
- use the `scale_...` functions to adjust colors, or any aesthetic

## using color palettes within ggplot
- default
```{r}
ggplot(iris)  + 
  geom_point(aes(x=Sepal.Length, Petal.Length, color = Sepal.Width))+
  theme_classic()
```

## using color palettes within ggplot

```{r}
ggplot(iris)  + 
  geom_point(aes(x=Sepal.Length, Petal.Length, color = Sepal.Width))+
  scale_color_gradientn(colors = pals::brewer.reds(100))+
  theme_classic()
```

## the `scale_...` functions 
- all `scale_...` functions have roughly the same format "scale_<aesthetic>_method"

```{r}
pal = RColorBrewer::brewer.pal(3, 'Pastel1')
ggplot(iris)  + 
  geom_boxplot(aes(x=Species, y=Petal.Length, fill = Species))+
  scale_fill_discrete(type = pal)+
  theme_classic()
```

## quick note about using discrete scales from the `pals` package
- `pals` returns named vectors, but are incompatible with the `scale_*_discrete` functions. Remember to remove the names from 

```{r}
pal = pals::alphabet()
names(pal) <- NULL
ggplot(iris)  + 
  geom_boxplot(aes(x=Species, y=Petal.Length, fill = Species))+
  scale_fill_discrete(type = pal)+
  theme_classic()
```

## the `scale_...` functions 
- always remember to set the right aesthetic 

```{r}
pal = RColorBrewer::brewer.pal(3, 'Pastel1')
ggplot(iris)  + 
  geom_boxplot(aes(x=Species, y=Petal.Length, fill = Species))+
  scale_color_discrete(type = pal)+
  theme_classic()
```

## the `scale_...` functions 

```{r}
plot_pal(pal)
```

## the `scale_...` functions 

```{r}
ggplot(iris)  + 
  geom_boxplot(aes(x=Species, y=Petal.Length, fill = Species))+
  scale_fill_discrete(type =pal)+
  theme_classic()
```


## the `scale_*_manual` functions 
- a subset of the `scale` functions, these allow you to manually specify a mapping between an aesthetic and value being plotted, for a discrete scale 

```{r}
col_vec <- c('setosa' = 'blue','versicolor' = 'purple',  'virginica' = 'red')
ggplot(iris) + 
  geom_boxplot(aes(x= Species, y= Sepal.Width, fill = Species)) +
  scale_fill_manual(values = col_vec)  + 
  theme_classic()
```

## the `scale_*_manual` functions 
-  using `scale_*_manual` is the best way to use a discrete color scale


## other aesthetics 
- there are a couple other aesthetics we haven't covered; some are specific to different plots 

## size
- size is pretty straight forward, bigger number > bigger size 

```{r}
ggplot(iris , aes(x=Sepal.Length, y=Sepal.Width, color = Species)) + 
  geom_point(size = 1)
```

## size

```{r}
ggplot(iris , aes(x=Sepal.Length, y=Sepal.Width, color = Species)) + 
  geom_point(size = 3)
```

## alpha
- alpha controls the transparency of geoms, and is between 0(translucent) and 1(solid fill)

```{r}
ggplot(iris , aes(x=Sepal.Length, y=Sepal.Width, color = Species)) + 
  geom_point(size = 3, alpha = .5)
```

## width
-  width is an example of an aesthetic specific to a few geoms `geom_boxplot`, `geom_bar`, and `geom_boxplot`, to name a few 

```{r}
ggplot(iris) + 
  geom_boxplot(aes(x= Species, y= Sepal.Width, fill = Species), width = .1) +
  scale_fill_manual(values = col_vec)  + 
  theme_classic()
```

## width
- always a good idea to spend some time looking through the documentation about what aesthetics work for what geoms 


