---
title: "R Notebook"
output: html_notebook
---

# Practice with `ggplot2`

We're going to go through several of the common geoms in ggplot. If at anytime you are stuck, try using the `?` to get the documentation for a function , or googling the problem. 

We're making plots in Rmarkdown; plots will display below the chunk once you run the chunk with the play button. 

You may see a prompt in the console asking you to select CRAN mirror. If you do, type 1 and press enter

```{r example_1}
install.packages('ggplot2')
```

load the library

```{r example_2}
library(ggplot2)
```

## `geom_point`

First, we'll be going through the geom_point function. We'll be using the `CO2` dataframe, The CO2 data frame has 84 rows and 5 columns of data from an experiment on the cold tolerance of the grass species Echinochloa crus-galli.

```{r ex_1}
data("CO2")
CO2
```

load the ggplot2 library, and make a scatter plot using `conc` for the x axis and `uptake` for the y axis.

```{r ex_2}

```


Now change all the points to the color red

```{r ex_3}

```


Change the color of the points to color by the `Type` column

```{r ex_4}

```

Change the color of the points to color by the `Plant` column

```{r ex_5}

```


note that this coloring is a continuous coloring, ie a uniform transition between shades of color. Continuous data types like `numeric` or `ordered.factor` are automatically given continuous scales, whereas categorical types like `character` or `factor` are colored with a discrete scales. We'll talk more about this when we discuss colors 

change the `Plant` column to a character and re create the plot above to see the change in color

```{r ex_6}

```

- now change the shape of the points based on the `Treatment` column.  A combination of shapes and colors is a good way to show two different variables

```{r ex_7}


```

the `alpha` value of a `geom` sets how transparent it is, with 1 being fully visible, and 0 being fully transparent

set the `alpha` value for all points to .75, then .5, then .25 to see how the plot changes

```{r ex_8}

```


## line graphs 

We'll be working with the `Indometh` data set, which has data related  Pharmacokinetics of Indomethacin
```{r example_3}
#run me
data("Indometh")
Indometh
```

make a basic line graph comparing `time` and `conc` from the `Indometh` data 

```{r ex_9}

```


Now make separate lines by coloring based on the `Subject` column
```{r ex_10}

```

use `as.character` to convert the `Subject` column to a character, and remake the same plot as above

```{r ex_11}

```

instead of connecting the data, use fitted line to make a similar plot as above, with separate lines for each subject. Try out a couple different smoothing methods

```{r ex_12}

```


## Bargraphs 

we're going to practice making bar graphs use the `mpg` data set which contains Fuel economy data from 1999 and 2008 for 38 popular models of car

```{r example_4}
data(mpg)
mpg
```

make a bargraph for  the `manufacturer` column

```{r ex_13}
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer))

```

- depending on the size of your screen, the labels on the x-axis maybe overlapping. We'll learn how to fix this in a later lesson

plot a stacked bar chart using `class` as the coloring variable. 

```{r ex_14}

```

now plot a bargraph where the Y axis shows the `hwy` column, and `class` for the x axis

```{r ex_15}


```

convert the `cyl` column to a `character`, and then make a dodged barplot using `cyl` as the coloring variable 

```{r ex_16}

```


## Boxplots 
 
We'll be working with the ToothGrowth data set, which has data related to the effect of Vitamin C on Tooth Growth in Guinea Pigs
 
```{r example_5}
data("ToothGrowth")
ToothGrowth
```

use a boxplot to visualize the length(`len`) distributions for the different treatment conditions(`supp`). Use the `?` command if you get stuck

```{r ex_17}


```


add `theme_minimal` to this plot if you have not done so; note the changes. 
Try out these other theme presets below:
`theme_bw()`
`theme_minimal()`
`theme_classic()`

```{r ex_18}

```


change the shading(fill) of the box to fill the box based on treatment
```{r ex_19}

```


change the width of all boxplots 
```{r ex_20}

```


## Visualization distributions 

Below is some fake data about the ages of patients in various part of a hospital. We are trying to visualize whether the distributions of ages in each region are different

When running the chunk below, a prompt in the console may appear and asj if you want to recompile the package. Type the NO option

```{r example_6}
install.packages('tidyverse')
library(tidyverse)
mu <- 2
si <- 0.6
hospital_data <- data.frame( ICU  =  c(rnorm(1000, -mu, si), rnorm(1000, mu, si)), 
                         ER  = runif(1000, -4, 4),
                         Clinic   =  rnorm(2000, 0, si*4)) %>% 
    tidyr::pivot_longer(cols = ICU:Clinic, 'Hospital_location') %>% rename(age = 'value')

```

first, lets plot a boxplot; make multiple boxplots shaded by location

```{r ex_21}

```

now plot make violin plots for each location; take note of the distributions 

```{r ex_22}

```

now plot histograms, again shaded by location

```{r ex_23}


```

`geom_freqpoly` is  drop-in replacement for `geom_histogram`: instead of drawing bar's for the count in each bin, it draws points at each count, and then draws a line through each point. Try it out with the above data, tweaking the coloring if necessary

```{r ex_24}

```


## Practice with themes 

We're going to work with one of the plots we made earlier 


```{r example_7}
ggplot(mpg) + 
  geom_bar(aes(x=manufacturer))

```


Assign descriptive axis labels and a plot title to the plot

```{r ex_25}

```


fix the x axis labels similar to how it was done in the example. You will need to set the relevant parameter in theme using the `element_text` function, changing the `hjust` and angle params accoridngly. run ?element_text for extra help

```{r ex_26}

```


change the y axis title so it is parallel to the x axis title 

```{r ex_27}

```

change the font of the title to Palatino, the color to red, and make it italizced


```{r ex_28}

```






