---
title: "R Notebook"
output: html_notebook
---

## install complexHeatmap
```{r}
install.packages(c('BiocManager','viridisLite' ))
BiocManager::install('ComplexHeatmap')
library(ComplexHeatmap) 
library(tidyverse)
```

- here is some example data we will be using

```{r}
gene_expression_matrix <- readRDS('../src/gene_expression_ss.RDS')

```

## A single heatmap 

- plot a heat map of the data 

```{r ex_1}

```

- set the colors of the heat map using `colorRamp2`

```{r ex_2}
library(circlize)

```

- set the color to the viridis palette. you can generate a vector of viridis colors by using `viridisLite::viridis`

```{r ex_3}



```

- change the heatmap such that there is a white gap between each tile.

```{r ex_4}

```

- create row and column titles accordingly 

```{r ex_5}

```


- only cluster the columns

```{r ex_6}


```

## 2.7 Heatmap split

- split the heatmap into 3 subpieces, by column

```{r ex_7}




```


## Heatmap annotations 
- this file contains metadata about the samples in gene expression data. sample_accession corresponds to the name of each sample in the gene expression matrix.

```{r ex_8}
sample_metadata <- read_tsv('../src/gene_expression_sample_tsv.gz')

```

- create a simple column annotation that corresponds to the `Tissue` column in the sample metadata

```{r ex_9}


```



