---
title: "R Notebook"
output: html_notebook
---

## install complexHeatmap
```{r}
install.packages(c('BiocManager','viridisLite' ))
BiocManager::install('ComplexHeatmap')
library(ComplexHeatmap) 
library(tidyverse)
```

- here is some example data we will be using
- log transforming gene expression data is commonly done before plotting.
```{r}

gene_expression_matrix <- readRDS('../src/gene_expression_ss.RDS')

```

## A single heatmap 

- plot a heat map of the data 
```{r}
Heatmap(gene_expression_matrix)
```

- set the colors of the heat map using `colorRamp2`

```{r}
library(circlize)
col_fun = colorRamp2(c(0, 50, 100), c("purple", "green", "yellow"))
col_fun(seq(-3, 3))
Heatmap(gene_expression_matrix, name = "mat", col = col_fun)
```

- set the color to the viridis palette. you can generate a vector of viridis colors by using `viridisLite::viridis`

```{r}
Heatmap(gene_expression_matrix, name = "mat", col = viridisLite::viridis(100))


```

- change the heatmap such that there is a white gap between each tile.

```{r}
Heatmap(gene_expression_matrix, name = "mat", col = viridisLite::viridis(100), 
        rect_gp = gpar(col = "white", lwd = 5))
```

- create row and column titles accordingly 

```{r}
Heatmap(gene_expression_matrix, name = "mat", col = viridisLite::viridis(100), 
        rect_gp = gpar(col = "white", lwd = 1), 
        column_title = 'Samples', row_title = 'genes')
```


- only cluster the columns

```{r}

Heatmap(gene_expression_matrix, name = "mat", col = viridisLite::viridis(100), 
        rect_gp = gpar(col = "white", lwd = 1), 
        column_title = 'Samples', row_title = 'genes', 
        cluster_rows = F)

```

## 2.7 Heatmap split

- split the heatmap into 3 subpieces, by column

```{r}


Heatmap(gene_expression_matrix, name = "mat", col = viridisLite::viridis(100), 
        rect_gp = gpar(col = "white", lwd = 1), 
        column_title = 'Samples', row_title = 'genes', 
        cluster_rows = F, 
        column_km = 3)

```


## Heatmap annotations 
- this file contains metadata about the samples in gene expression data. sample_accession corresponds to the name of each sample in the gene expression matrix.

```{r}

sample_metadata <- read_tsv('../src/gene_expression_sample_tsv.gz')

```

- create a simple column annotation that corresponds to the `Tissue` column in the sample metadata

```{r}

ha <- HeatmapAnnotation(Tissue = sample_metadata$Tissue)

Heatmap(gene_expression_matrix[,sample_metadata$sample_accession], 
        name = "mat", 
        col = viridisLite::viridis(100), 
        rect_gp = gpar(col = "white", lwd = 1), 
        column_title = 'Samples', row_title = 'genes', 
        cluster_rows = F, 
        column_km = 3, 
        top_annotation = ha)



```



