---
title: "R Notebook"
output: html_notebook
---

Here are come plots to work with 

## practice with simple plots 

```{r}
library(ggplot2)
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp,)) + 
  ggtitle('Plot 1')

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear, fill=as.character(gear))) + 
  ggtitle('Plot 2') + 
    guides(fill = guide_legend(title = 'gear'))

p3 <- ggplot(mtcars) + 
  geom_point(aes(hp, wt, colour = mpg)) + 
  ggtitle('Plot 3')

p4 <- ggplot(mtcars) + 
  geom_bar(aes(gear)) + 
  facet_wrap(~cyl) + 
  ggtitle('Plot 4')
```

- add the plots together one by one, no orientation

```{r ex_1}

```


- You can add the parameters `fig.width` and `fig.height` in the R chunk to change the plot output dimensions. An example of this is shown in the chunk below
- add plots, but make only one column of plots. 

```{r ex_2, fig.width=4, fig.height=10}

```

- use a design vector and `plot_layout` to make plot 1 bigger than the rest of the plots 

```{r ex_3}

```


```{r ex_4}
'../Advanced-Plotting-Libraries/useful_vis_exercises.Rmd'
```


## A slightly more realistic example

- we have made multiple plots using an scRNA-seq data set ('../src/pbmc_cell_expression.tsv.gz'). We are going to make a couple new plot, and  re-create previous ones we made and combine all of them together using patchwork

For each of the plots, go to the notebook we worked on the plots and bring the code into here. Store each plot in a separate variable.
Steps
- load the data in. 
- make a scatterplot using the UMAP columns in the data as the axes. Color the plot by celltype 
- For all the plots below, log transform( data_log = log2(data + 1)) before plotting. 
- make a dotplot like we did in '../Plotting-With-ggplot/ggplot2_basics_exercises.Rmd', transforming the data accordingly
- make a Heatmap using `ComplexHeatmap`. Use the code from '../Advanced-Plotting-Libraries/learning_new_packages.Rmd' as a guide 
  - you will need to use the `ggplotify` library to convert this heatmap to ggplot. Make sure you install and load the package before using it
- make a raincloud plot for a gene(colored by cell type) like we did in '../Advanced-Plotting-Libraries/useful_vis_exercises.Rmd'
- Create a color mapping between celltype and colors, so you can keep colors consistent across multiple plots. Use this along with `scale_*_manual`
- clean up axis titles and legend labels( ie no underscores)
- combine these all together with patchwork
  - create a design vector so all plots are equally visible. 

```{r ex_5}

```





